---
title: "R tidyverse basics"
output: html_notebook
---

The Tidyverse is “an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures.”

```{r}
library(tidyverse)
setwd("/path/to/your/working/directory")
```


### TIDY DATA ###
Key to the Tidyverse system is the tidy data format, where:

* Each variable must have its own column.
* Each observation must have its own row.
* Each value must have its own cell.

Let's read in some data.
```{r}
eagle_nest_data <- data.frame(
  "region" = c("Pacific", "Southwest", "Rockies"),
  "2007" = c(1039, 51, 200),
  "2009" = c(2587, 176, 338),
  check.names=FALSE
)

print(eagle_nest_data)
```
The variables in this dataset are region, year, and nests.

* Region is tidy because all of the data is in a single column. 
* Year is not tidy because the values (2007, 2009) are in two different columns.
* Nests is not tidy because the values are stored as values of 2007 and 2009.

Now we'll make the data tidy with the pivot_longer( ) function.
```{r}
eagle_nest_data %>%           # %>% (magrittr) is a pipe (forwards a value into an expression or function).
  pivot_longer(
    cols = c(`2007`, `2009`), # Reorganize so that 2007 and 2009 are in a single column.
    names_to = "year",        # Rename that column "year".
    values_to = "num_nests"   # 2007 and 2009 nest values are in a new "num_nests" column.
  )
```
Now each variable has a column and each data point has its own row.

For a more thorough explanation see https://dcl-wrangle.stanford.edu/pivot-basic.html


### IMPORTING DATA ###
We will use the read_csv( ) function from the readr package to upload global tuberculosis data as a dataframe. Learn more about the data here: https://tidyr.tidyverse.org/reference/who.html
```{r}
TBdata <- read_csv("/path/to/TBdata_tidy.csv")
```

### GROUPING DATA ###
The package dplyr allows you to group data by variable(s). 

To get an idea of the data, let's look at the first 6 rows. <br>
Age values are a range (eg., 3544 = ages 35 to 44). <br>
Var refers to the method of diagnosis (rel = relapse, sn = negative pulmonary smear, sp = positive pulmonary smear, ep = extrapulmonary).
```{r}
head(TBdata)
```

To simplify analysis, let's pull out the data for one country.
```{r}
brazilTB <- filter(TBdata, grepl('Brazil', TBdata$country))
```

The data can be grouped by any variable(s).
```{r}
# Group by 1 variable
brazilTB_by_sex <- brazilTB %>% group_by(sex)

# Group by multiple variables
brazilTB_by_sex_var <- brazilTB %>% group_by(sex, var)
```

Let's count the number of rows (i.e., data points) in each group.
```{r}
brazilTB_by_sex %>% tally()
```

Or sort the number of rows from largest to smallest.
```{r}
brazilTB_by_sex_var %>% tally(sort=TRUE)
```

We can get a group's basic statistics with the summarize( ) function.
```{r}
brazilTB_by_sex_var %>% summarize(mean_cases = mean(cases),  max_cases=max(cases))
```


### VISUALIZING DATA ###
The Tidyverse package for visualizing data is ggplot2.

Let's load the gapminder data set, which is a global database of various metrics.
```{r}
if(!require(gapminder)) {install.packages("gapminder"); library(gapminder)}
```

What are the variables of this version of the gapminder data?
```{r}
str(gapminder)
```
Plot (graph) GDP over time. Each point represents a country.
```{r}
ggplot(data = gapminder, aes(x = year, y = gdpPercap)) + 
    geom_point()
```

We can also combine plots.
```{r}
ggplot(gapminder, aes(x = year, y = lifeExp, group = country)) +
  geom_line(aes(color = continent)) +
  geom_point(aes(size = gdpPercap))
```

We can subset the data and plot each subset separately with facets.
```{r}
ggplot(gapminder, aes(x = year, y = lifeExp, group = country)) +
  geom_line(aes(color = continent)) +
  facet_grid(~ continent)
```

Finally, let's change the axis labels to make the plot more readable.
```{r}
ggplot(gapminder, aes(x = year, y = lifeExp, group = country)) +
  geom_line(aes(color = continent)) +
  facet_grid(~ continent) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  ylab("Life Expectancy") +
  xlab("Year") +
  theme(legend.title=element_blank())

```